### 2021-10-26 documents/sections/extensions.tex
rewriting of extensions section

\section{The innovative \gl{1}\gl{s_a_} markers}
\label{sec:extensions}
As stated in \cref{sec:extensions_intro}, the person marker extensions which did not spread through the entire lexicon all have in common that they feature innovative \gl{1}\gl{s_a_} markers.
There are six attested such innovations, three of which can be reconstructed to intermediate proto-languages.
The new \gl{1}\gl{s_a_} prefix is formally identical to the \gl{1+2} marker (\PC \rc{k-}) in three cases, to the \gl{1}\gl{s_p_} marker (\PC \rc{u(j)-}) in two cases, and to the \gl{1}>\gl{3} marker (\PC \rc{t-}) in one case.
\cref{sec:pekodian} investigates the innovation of \rc{k-} in \PPek, reflected in the three daughter languages \arara, \ikpeng, and \bakairi.
\cref{sec:waiwaian} takes a closer look at the extension of \rc{k-} in \PWai, which was briefly shown in \cref{sec:extensions_intro}.
\cref{sec:taranoan} concerns the extension of \rc{t-} in \PTir or \PTar, reflected in modern \trio and \akuriyo.
\crefrange{sec:akuriyo}{sec:yukpa} look at innovative first person markers which are only attested in single modern languages:
\obj{k-} in \akuriyo, and \obj{j-} in \carijo and \yukpa.

@@

For \arara, \textcite{alves2017arara} reports six\footnote{Seven under her analysis, which treats the two meanings of \obj{it͡ʃi} \qu{to be, to lie down} as different verbs.} \gl{s_a_} verbs which have a first person marker \obj{w(ɨ)-} rather than \obj{k-}, listed in \exref{arairr}.
%We take these \arara verbs as a starting point for the other Pekodian languages, as \textcite{alves2017arara} gives the most detailed description of person markers for any language of the branch.

\ex<arairr> \arara \parencite[153]{alves2017arara} \\
\begin{tabular}[t]{@{}ll@{}}
\obj{wɨ-genɨ} & \qu{I said}\\
\obj{w-it͡ʃinɨ} & \qu{I was, lied down}\\
\obj{w-ebɨnɨ} & \qu{I came}\\
\obj{w-ibɨnɨ} & \qu{I bathed}\\
\obj{w-iptoŋrɨ} & \qu{I went down}\\
\obj{w-ɨdolɨ} & \qu{I went}\\
\end{tabular}
\xe
%
From a comparative view, this list is not quite complete, as there is also a reflex of the copula \rc{a[p]}, which is analayzed as a postposition by \textcite[199--201]{alves2017arara}; it is however inflected with \setone prefixes and also shows first person \obj{w-} \exref{araap}.

\ex<araap>
\begin{tabular}[t]{@{}ll@{}}
\gl{1} & \obj{w-aptam}\\
\gl{2} & \obj{m-od-aptam}\\
\gl{1+2} & \obj{kud-aptam}\\
\gl{3} & ∅\obj{-aptam}\\
\end{tabular}
\xe

@@

\ex<bak-33>\bakairi \parencite[][4]{meira2003bakairi}\\
\begingl
\gla w-i-də//
\glb \gl{1}\gl{s_a_}-bathe-\gl{imm}//
\glft \qu{I bathed}//
\endgl
\xe
%
While \textcite[4]{meira2003bakairi} does list \obj{ge} \qu{to say}, \obj{tə} \qu{to go}, and \obj{əe(wɨ)} \qu{to come} as examples of \gl{s_a_} verbs, he does not indicate whether they belong to the class of \gl{s_a_}-1 verbs, with first person \obj{k-}, or the \gl{s_a_}-2 verbs, with \obj{w-}.\footnote{It should be noted at this point that \textcite{meira2003bakairi} indicates that the same verbs which take first person \obj{w-} in \bakairi also take a \gl{1+2} marker \obj{k-}. However, this marker is only illustrated for \qu{to bathe}, both by \textcite{meira2003bakairi} and \textcite{von1892bakairi}. Given the lack of data for other verbs, we will not further discuss this potential additional pattern. If the characterization by \citeauthor{meira2003bakairi} is accurate, then the pattern is fully parallel to the distribution of the first person prefixes.}
Luckily, while \textcite{von1892bakairi} did not accurately record all phonemic distinctions in \bakairi \parencite{meira2005bakairi}, he does provide inflected forms of cognates to the \arara verbs in \exref{arairr}.
We present them according to the phonemic analysis of \bakairi by \textcites{wheatley1969bakairi}{meira2003bakairi}{meira2005bakairi} in \exref{bakverbs}.

@@

All available descriptions of the third Pekodian language, \ikpeng, list \obj{k-} as the only \gl{1}\gl{s_a_} marker \parencites[55]{ikpengpacheco1997}[105]{campetela1997analise}[64]{ikpengpacheco2001}[205]{alves2013verbo}.
However, most \ikpeng cognates of the \arara verbs with \gl{1}\gl{s_a_} \obj{w-} actually do not take \obj{k-}, but rather \obj{ɨ-} or Ø, as shown in \exref{ikpw}.
The exception is \qu{to go}, which has \obj{k-} \exref{ikp-170}.
There is a formally identical \ikpeng cognate of \arara \obj{iptoŋ} \qu{to go down}, but no first person forms are attested \perscommpar{Angela Chagas}.
While there are reflexes of \rc{a[p]} \qu{to be} in \ikpeng, only \rc{eti} forms are attested with first person inflectional prefixes \parencite[401]{gildea2018reconstructing}.

@@

\ex<ikp-170>\ikpeng \parencite[][80]{ikpengpacheco2001}\\
\begingl
\gla k-aran-t͡ʃi//
\glb \gl{1}-go-\gl{npst}//
\glft \qu{I'm going.}//
\endgl
\xe

%\ex<ikp-182>\ikpeng \parencite[][320]{chagas2019ikpeng}\\
%\begingl
%\gla ∅-iptoŋ-lan//
%\glb \gl{3}\gl{s_a_}-sit.down-\gl{pst}//
%\glft \qu{She sat down.}//
%\endgl
%\xe
%
While clearly not a regular sound change, \PXin \rc{w} is attested as sometimes being lost in \ikpeng, as evidenced by the correspondences in \cref{tab:pxinw}.
Thus, we suggest that the \ikpeng \gl{1}\gl{s_a_} marker \obj{ɨ-}/Ø is cognate with \arara \gl{1}\gl{s_a_} \obj{w(ɨ)-}, and was subject to phonological erosion of \rc{w}, a phenomenon which will also be seen in other languages discussed in following sections.

@@

@@

\Cref{tab:ppekverbs} gives an overview of the first person forms of the seven verbs under discussion, along with our \PPek reconstruction.
The form for \rc{ge} \qu{to say} is straightforward to reconstruct; the \bakairi change of \rc{wɨ-} to \obj{u-} is not unexpected given correspondences like \bakairi \obj{udo} \parencite{meira2005southern} from \PC \rc{wɨtoto} \qu{person} \parencite[4]{gildea2007greenberg}.
\qu{to bathe} is straightforwardly reconstructible as \rc{ipɨ}.
For \qu{to be}, \ikpeng \obj{e} is very likely the original vowel, given the \PC form \rc{eti} \pcref{sec:be}.

The forms for \qu{to come} are not fully cognate; \ikpeng and \bakairi both show a reflex of the detransitivizer in combination with a root reconstructible as \rc{epɨ}.
In contrast, the \arara first person form is directly based on this root \rc{epɨ}.
However, other person values show a reflex of \detrz in \arara, too \exref{ara-123}.

\ex<ara-123>\arara \parencite[][150]{alves2017arara}\\
\begingl
\gla m-odebɨ-nɨ//
\glb \gl{2}\gl{s_a_}-come-\gl{rec}//
\glft \qu{You came.}//
\endgl
\xe
%
On the other hand, both \ikpeng and \bakairi show reflexes of \rc{əd-ebɨ} throughout the whole paradigm.
Following the line of reasoning used by \textcite[114]{meira1998proto} (see also \cref{sec:taranoan}) for a similar pattern in the three Taranoan languages, we suggest that the idiosyncratic pattern in \arara is reconstructible to \PPek, and that \bakairi and \ikpeng independently regularized the paradigm.

The forms for \qu{to go down} are also not fully cognate; \textcite{meira2005southern} make no mention of a regular correspondence between \bakairi \obj{gɨ} and \ikpeng \obj{ŋ}.
Rather, it seems that some roots gained a final \obj{ŋ} in \PXin, for so far unclear reasons: \begin{inlinelist}
\item \PC \rc{əne} \qu{to see}, \arara and \ikpeng \obj{eneŋ} \parencites[8]{gildea2007greenberg}[56]{alves2017arara}[25]{ikpengpacheco2001}
\item \PC \rc{əta} \qu{to hear}, \arara \obj{taŋ}, \ikpeng \obj{iraŋ} \parencites[8]{gildea2007greenberg}[144]{alves2017arara}[270]{ikpengpacheco2001}
\item \PC \rc{ənə} \qu{to eat meat}, \arara \obj{oŋoŋ} \qu{to bite} \parencites[8]{gildea2007greenberg}[57]{alves2017arara}
\end{inlinelist}.
Also, based on the fact \bakairi has generally lost much segmental material, we suggest that \bakairi \obj{gɨ} is yet another addition to a root \rc{ɨptə} or \rc{iptə}, rather than a conservative form.

The forms for \qu{to go} are all cognate, but while \arara and \bakairi point to \rc{ɨtə}, the \ikpeng form seems rather divergent.
However, it is in fact compatible with a reconstruction \rc{ɨtən}, when one considers that \ikpeng \obj{a} is an attested outcome of \rc{ə} (\PXin \rc{o}/\obj{e}), for example \obj{akari} \qu{dog} in \exref{ikpwloss} above, \obj{anma} \qu{path} \parencite[24]{ikpengpacheco2001} from \PC \rc{ətema} \parencite[12]{gildea2007greenberg}, or \obj{jaj} \qu{tree} from \PC \rc{jəje}.
This attested change of \rc{ə} to \obj{a} need only be preceded by a assimilatory lowering of initial \rc{ɨ} to \rc{ə}, to yield the form \obj{aran} from \rc{ɨtən}.
Other \ikpeng reflexes of \qu{to go} offer evidence for the suggested intermediate stage \rc{ətən}: \obj{ero-lɨ} \qu{s/he went} \parencite[25]{ikpengpacheco2001}.

Summing up, an innovative \gl{1}\gl{s_a_} marker \rc{k-} is reconstructible to \PPek.
Seven verbs can be reconstructed as having resisted this innovation, and preserving \gl{1}\gl{s_a_} \rc{w(ɨ)-} in \PPek.
For \qu{to come}, two distinct forms \rc{w-epɨ-} and \rc{k-əd-epɨ-} can be reconstructed, although the latter may be two parallel developments in \bakairi and \ikpeng.
In later, individual developments, \bakairi introduced \obj{k-} to \qu{to go down}, and \ikpeng to \qu{to go}.

@@

\ex<wais> Waiwaian \rc{eɸurka} \qu{to fall} (\gl{s_a_}) and \rc{wɨnɨkɨ} \qu{to sleep} (\gl{s_p_})\\
\begin{tabular}[t]{@{}lllllll@{}}
& \multicolumn{2}{l}{\PWai} & \multicolumn{2}{l}{\hixka}  & \multicolumn{2}{l}{\waiwai} \\
 & \gl{s_a_} & \gl{s_p_} & \gl{s_a_} & \gl{s_p_} & \multicolumn{2}{l}{\gl{s}} \\
\gl{1} & \rc{\emp{k-}eɸurka-} & \rc{kɨ-wɨnɨkɨ-} & \obj{\emp{k-}ehurka-} & \obj{kɨ-nɨkɨ-} & \obj{\emp{k-}eɸɨrka-} & \obj{kɨ-wɨnk-} \\
\gl{2} & \rc{m-eɸurka-} & \rc{o-wɨnɨkɨ-} & \obj{m-ehurka-} & \obj{o-wnɨkɨ-} & \obj{m-eɸɨrka-} & \obj{mɨ-wɨnk-} \\
\gl{1+2} & \rc{t-eɸurka-} & \rc{tɨt-wɨnɨkɨ-} & \obj{t-ehurka-} & \obj{tɨ-nɨkɨ-} & \obj{t͡ʃ-eɸɨrka-} & \obj{tɨt-wɨnk-} \\
\gl{3} & \rc{ɲ-eɸurka-} & \rc{nɨ-wɨnɨkɨ-} & \obj{ɲ-ehurka-} & \obj{nɨ-nɨkɨ-} & \obj{ɲ-eɸɨrka-} & \obj{nɨ-wɨnk-} \\
\end{tabular}\\
\parencites[510]{howard2001wrought}[189--191]{hixkaryanaderby1985}[209--211]{hawkins1953waiwai}[50]{waiwaihawkins1998}
\xe
%
\waiwai \obj{ka} \qu{to say} does not take \obj{kɨ-}, but rather conservative \obj{wɨ-} \exref{hixɨ.wai-108}.
Its \hixka counterpart also takes a reflex of \gl{1}\gl{s_a_} \rc{w-}, which has the form \obj{ɨ-} \exref{hixɨ.hix-118}.
A formally identical prefix is found on transitive verbs in \gl{1}>\gl{3} scenarios \exref{hixɨ.hix-17}, where it also corresponds to \waiwai \obj{w(ɨ)-} \exref{hixɨ.wai-159}.

\pex<hixɨ>
\a<wai-108> \waiwai \parencite[][71]{waiwaihawkins1998}\\
\begingl
\glpreamble wɨɨkekɲe//
\gla wɨ-ka-jakɲe//
\glb \gl{1}-say-\gl{pst}//
\glft \qu{I said.}//
\endgl
\a<hix-118> \hixka \parencite[][124]{hixkaryanaderby1985}\\
\begingl
\glpreamble roxehra nay hamɨ Kaywerye ɨkekonɨ//
\gla ro-ʃe-hɨra n-a-je hamɨ kajwerʲe ɨ-ka-jakonɨ//
\glb \gl{1}-\gl{des}-\gl{neg} \gl{3}-be-\gl{npst}.\gl{uncert} \gl{evid} K. \gl{1}\gl{s_a_}-say-\gl{rem}.\gl{cont}//
\glft \qu{I said (to myself), “Kaywerye evidently doesn't like me”.}//
\endgl
\a<hix-17> \hixka \parencite[][191]{hixkaryanaderby1985}\\
\begingl
\gla ɨ-koroka-no//
\glb \gl{1}>\gl{3}-wash-\gl{imm}//
\glft \qu{I washed him.}//
\endgl
\a<wai-159> \waiwai \parencite[][192]{waiwaihawkins1998}\\
\begingl
\glpreamble wîyesî//
\gla wɨ-jo-jasɨ//
\glb \gl{1}>\gl{3}-boil-\gl{npst}//
\glft \qu{I will boil it.}//
\endgl
\xe
%
Thus, similar to the the case of \ikpeng, \hixka has lost the \rc{w} of the \envr{}{C} allomorph \rc{wɨ-}.
Notably, \textcite{hixkaryanaderby1985} does not see this \obj{ɨ-} as an irregular \gl{1}\gl{s_a_}, but as the regular \gl{1}>\gl{3} marker, because he considers \hixka \obj{ka} \qu{to say} to be transitive, an issue which will be discussed in \cref{sec:say}.

There are three more verbs which did not take innovative \rc{k-} in \PWai; they are shown along with \rc{ka} \qu{to say} in \cref{tab:pwaiverbs}.
While the two roots for \qu{to be} are straightforwardly reconstructible, \qu{to go} is somewhat of a special case.
While \hixka has the expected \obj{ɨ-}, \waiwai seems to have combined innovative \gl{1}\gl{s_a_} \obj{k-} with the old \rc{w-}, an analysis also considered by \textcite[90]{gildea1998}.
Alternatively, this form was influenced by deverbalized forms of \qu{to go}, where a reflex of the \gl{s_a_} class marker \rc{w-} has become fossilized \exref{waiwto}.

\pex<waiwto> \waiwai reflexes of the \gl{s_a_} class marker \rc{w-}
\a \obj{o-\emp{w}to-topo-nho} \qu{my trip} \parencite[92]{waiwaihawkins1998}
\a \obj{o-\emp{w}to-t͡ʃhe} \qu{after I went} \parencite[165]{waiwaihawkins1998}
\a \obj{kɨ-\emp{w}to-me} \qu{for us to go} \parencite[204]{waiwaihawkins1998}
\xe
%
In any case, \hixka \qu{to go} was clearly not affected by the extension of \rc{k-}, allowing us to reconstruct a \PWai first person form \rc{wɨ-to-}.

@@

Summing up, we reconstruct the three verbs \rc{et͡ʃi}/\obj{a(h)} \qu{to be}, \rc{ka(s)} \qu{to say}, and \rc{to} \qu{to go} as preserving the old \gl{1}\gl{s_a_} marker \rc{w-} in \PWai, while the rest took on innovative \rc{k-}.

>>>>

\section{The innovative \gl{1}\gl{s_a_} markers}
\label{sec:extensions}
As stated in \cref{sec:extensions_intro}, the person marker extensions which did not spread through the entire lexicon all have in common that they feature innovative \gl{1}\gl{s_a_} markers.
There are six attested such innovations, three of which can be reconstructed to intermediate proto-languages.
The new \gl{1}\gl{s_a_} prefix is formally identical to the \gl{1+2} marker (\PC \rc{k-}) in three cases, to the \gl{1}\gl{s_p_} marker (\PC \rc{u(j)-}) in two cases, and to the \gl{1}>\gl{3} marker (\PC \rc{t-}) in one case.
\cref{sec:pekodian} investigates the innovation of \rc{k-} in \PPek, reflected in the three daughter languages \arara, \ikpeng, and \bakairi.
\cref{sec:waiwaian} takes a closer look at the extension of \rc{k-} in \PWai, which was briefly shown in \cref{sec:extensions_intro}.
\cref{sec:taranoan} concerns the extension of \rc{t-} in \PTir (or \PTar), reflected in modern \trio and \akuriyo.
\crefrange{sec:akuriyo}{sec:yukpa} look at innovative first person markers which are only attested in single modern languages:
\obj{k-} in \akuriyo, and \obj{j-} in \carijo and \yukpa.

@@

For \arara, \textcite{alves2017arara} describes six\footnote{Seven under her analysis, which sees the two meanings of \obj{it͡ʃi} \qu{to be, to lie down} as different verbs.} \gl{s_a_} verbs which have a first person marker \obj{w(ɨ)-} rather than \obj{k-}, listed in \exref{arairr}.
From a comparative view this list is not quite complete, as there is also a reflex of the copula \rc{a[p]}, serving syntactically as a postposition introducing adverbial clauses meaning \qu{if} or \qu{when} \parencite[199--201]{alves2017arara}.
However, its inflectional morphology features verbal \setone prefixes, including first person \obj{w-} \exref{araap}.
%We take these \arara verbs as a starting point for the other Pekodian languages, as \textcite{alves2017arara} gives the most detailed description of person markers for any language of the branch.

\begin{multicols}{2}
\ex<arairr> \arara \parencite[153]{alves2017arara} \\
\begin{tabular}[t]{@{}ll@{}}
\obj{wɨ-genɨ} & \qu{I said}\\
\obj{w-it͡ʃinɨ} & \qu{I was, lied down}\\
\obj{w-ebɨnɨ} & \qu{I came}\\
\obj{w-ibɨnɨ} & \qu{I bathed}\\
\obj{w-iptoŋrɨ} & \qu{I went down}\\
\obj{w-ɨdolɨ} & \qu{I went}\\
\end{tabular}
\xe
%
\ex<araap> \arara \parencite[200]{alves2017arara}\\
\begin{tabular}[t]{@{}ll@{}}
\gl{1} & \obj{w-aptam}\\
\gl{2} & \obj{m-od-aptam}\\
\gl{1+2} & \obj{kud-aptam}\\
\gl{3} & ∅\obj{-aptam}\\
\end{tabular}
\xe
\end{multicols}

@@

\ex<bak-33>\bakairi \parencite[][4]{meira2003bakairi}\\
\begingl
\gla w-i-də//
\glb \gl{1}\gl{s_a_}-bathe-\gl{imm}//
\glft \qu{I bathed}//
\endgl
\xe
%
While \textcite[4]{meira2003bakairi} does list \obj{ge} \qu{to say}, \obj{tə} \qu{to go}, and \obj{əe(wɨ)} \qu{to come} as examples of \gl{s_a_} verbs, he does not indicate whether they belong to the class of \gl{s_a_}-1 verbs, with first person \obj{k-}, or the \gl{s_a_}-2 verbs, with \obj{w-}.\footnote{It should be noted at this point that \textcite{meira2003bakairi} indicates that the same verbs which take first person \obj{w-} in \bakairi also take a \gl{1+2} marker \obj{k-}. However, this marker is only illustrated for \qu{to bathe}, both by \textcite{meira2003bakairi} and \textcite{von1892bakairi}. Given the lack of data for other verbs, we will not further discuss this potential additional pattern. If the characterization by \citeauthor{meira2003bakairi} is accurate, then the pattern is fully parallel to the distribution of the first person prefixes.}
Luckily, while \textcite{von1892bakairi} did not accurately record all phonemic distinctions in \bakairi \parencite{meira2005bakairi}, he does provide inflected forms of cognates to the \arara verbs in \exref{arairr}.
We present them according to our understanding of the phonemic analysis of \bakairi by \textcites{wheatley1969bakairi}{meira2003bakairi}{meira2005bakairi} in \exref{bakverbs}.

@@

All available descriptions of the third Pekodian language, \ikpeng, list \obj{k-} as the only \gl{1}\gl{s_a_} marker \parencites[55]{ikpengpacheco1997}[105]{campetela1997analise}[64]{ikpengpacheco2001}[205]{alves2013verbo}.
However, most \ikpeng cognates of the \arara verbs with \gl{1}\gl{s_a_} \obj{w-} actually do not take \obj{k-}, but rather \obj{ɨ-} or Ø, as shown in \exref{ikpw}.
The exception is \qu{to go}, which has \obj{k-} \exref{ikp-170}.
There is a formally identical \ikpeng cognate of \arara \obj{iptoŋ} \qu{to go down}, but no first person forms are attested \perscommpar{Angela Chagas}.
Further, while there are reflexes of \rc{a[p]} \qu{to be} in \ikpeng, only \rc{eti} forms are attested with first person inflectional prefixes \parencite[401]{gildea2018reconstructing}.

@@

\ex<ikp-170>\ikpeng \parencite[][80]{ikpengpacheco2001}\\
\begingl
\gla k-aran-t͡ʃi//
\glb \gl{1}-go-\gl{npst}//
\glft \qu{I'm going.}//
\endgl
\xe
%
%\ex<ikp-182>\ikpeng \parencite[][320]{chagas2019ikpeng}\\
%\begingl
%\gla ∅-iptoŋ-lan//
%\glb \gl{3}\gl{s_a_}-sit.down-\gl{pst}//
%\glft \qu{She sat down.}//
%\endgl
%\xe
%

@@

\Cref{tab:ppekverbs} gives an overview of the first person forms of the seven verbs under discussion, along with our \PPek reconstruction.
The presence of the \ikpeng \gl{1}\gl{s_a_} marker \obj{ɨ-}/Ø suggests to us that it may be cognate with \arara \gl{1}\gl{s_a_} \obj{w(ɨ)-}.
Indeed, \PXin \rc{w} is attested as sometimes being lost in \ikpeng, as evidenced by the correspondences in \cref{tab:pxinw}.
While it is by no means a regular sound change, it allows us to securely connect the two prefixes.
Similarly, the change of \rc{wɨ} to \bakairi \obj{u} is found in other correspondences, like \obj{udo} \parencite{meira2005southern} from \PC \rc{wɨtoto} \qu{person} \parencite[4]{gildea2007greenberg}.
Thus, we reconstruct a \gl{1}\gl{s_a_} prefix \rc{w(ɨ)-} to \PPek, identical to \arara.

As for the forms of the verb stems, a few comments are necessary:
For \qu{to be}, \ikpeng \obj{e} is very likely the original vowel, given the \PC form \rc{eti} \pcref{sec:be}.

For \qu{to go down}, we reconstruct \rc{ɨ} as the initial vowel rather than \rc{i}, see \cref{sec:godown}.
Further, the forms are not fully cognate; \textcite{meira2005southern} make no mention of a regular correspondence between \bakairi \obj{gɨ} and \ikpeng \obj{ŋ}.
However, the addition of a final \obj{ŋ} in \PXin is attested elsewhere: \begin{inlinelist}
\item \PC \rc{əne} \qu{to see}, \arara and \ikpeng \obj{eneŋ} %\parencites[8]{gildea2007greenberg}[56]{alves2017arara}[25]{ikpengpacheco2001}
\item \PC \rc{əta} \qu{to hear}, \arara \obj{taŋ}, \ikpeng \obj{iraŋ} %\parencites[8]{gildea2007greenberg}[144]{alves2017arara}[270]{ikpengpacheco2001}
\item \PC \rc{ənə} \qu{to eat meat}, \arara \obj{oŋoŋ} \qu{to bite} %\parencites[8]{gildea2007greenberg}[57]{alves2017arara}
\end{inlinelist} \parencites[8]{gildea2007greenberg}[56, 144, 57]{alves2017arara}[25, 270]{ikpengpacheco2001}.
Also, based on the fact \bakairi has generally lost much segmental material, we suggest that \bakairi \obj{gɨ} is another addition to a root \rc{ɨptə}, rather than a conservative form.

@@

The forms for \qu{to come} are not fully cognate, either: \ikpeng and \bakairi both show a reflex of the \PPek detransitivizer \rc{əd-} in combination with a root reconstructible as \rc{epɨ}.
In contrast, the \arara first person form is directly based on this root \rc{epɨ}.
However, other person values show a reflex of \rc{əd-} in \arara, too \exref{ara-123}.

\ex<ara-123>\arara \parencite[][150]{alves2017arara}\\
\begingl
\gla m-odebɨ-nɨ//
\glb \gl{2}\gl{s_a_}-come-\gl{rec}//
\glft \qu{You came.}//
\endgl
\xe
%
On the other hand, both \ikpeng and \bakairi show reflexes of \rc{əd-ebɨ} throughout the whole paradigm.
Following the line of reasoning used by \textcite[114]{meira1998proto} (see also \cref{sec:taranoan}) for a similar pattern in the three Taranoan languages, we suggest that the idiosyncratic pattern in \arara is reconstructible to \PPek, and that \bakairi and \ikpeng independently regularized the paradigm to only use \rc{əd-ebɨ}.

The V-initial nature of \qu{to go} is evidenced in its Xinguan forms, while \bakairi \phon{\rc{wɨ}}{\obj{u}} resulted in a shift of the morpheme boundary, see also \cref{sec:go}.
The \ikpeng form \obj{aran} is made compatible with our reconstruction \rc{ɨtən} by considering that \ikpeng \obj{a} is an attested outcome of \rc{ə} (\PXin \rc{o}/\obj{e}), for example \begin{inlinelist}
 \item \obj{akari} \qu{dog} in \cref{tab:pxinw} above
 \item \obj{anma} \qu{path} \parencite[24]{ikpengpacheco2001} from \PC \rc{ətema} \parencite[12]{gildea2007greenberg
 \item \obj{jaj} \qu{tree} \parencite[98]{ikpengpacheco2001} from \PC \rc{jəje}
 \end{inlinelist}.
This attested change of \rc{ə} to \obj{a} need only be preceded by a assimilatory lowering of initial \rc{ɨ} to \rc{ə}, to yield the form \obj{aran} from \rc{ɨtən}.
Other \ikpeng reflexes of \qu{to go} offer evidence for the suggested intermediate stage \rc{\underline{ə}t\underline{ə}n}: \obj{\underline{e}r\underline{o}-lɨ} \qu{s/he went} \parencite[25]{ikpengpacheco2001}.

Summing up, an innovative \gl{1}\gl{s_a_} marker \rc{k-} is reconstructible to \PPek.
Seven verbs can be reconstructed as having resisted this innovation, and preserving \gl{1}\gl{s_a_} \rc{w(ɨ)-} in \PPek.
In later, individual developments, \bakairi introduced \obj{k-} to \qu{to go down}, and \ikpeng to \qu{to go}.
Further, both languages regularized the paradigm of \qu{to come} to \rc{əd-epɨ} and introduced \obj{k-}.

@@

\ex<wais> Waiwaian \rc{eɸurka} \qu{to fall} (\gl{s_a_}) and \rc{wɨnɨkɨ} \qu{to sleep} (\gl{s_p_})\\
\begin{tabular}[t]{@{}lllllll@{}}
& \multicolumn{2}{l}{\PWai} & \multicolumn{2}{l}{\hixka}  & \multicolumn{2}{l}{\waiwai} \\
 & \gl{s_a_} & \gl{s_p_} & \gl{s_a_} & \gl{s_p_} & \multicolumn{2}{l}{\gl{s}} \\
\gl{1} & \rc{\emp{k-}eɸurka-} & \rc{kɨ-wɨnɨkɨ-} & \obj{\emp{k-}ehurka-} & \obj{kɨ-nɨkɨ-} & \obj{\emp{k-}eɸɨrka-} & \obj{kɨ-wɨnk-} \\
\gl{2} & \rc{m-eɸurka-} & \rc{o-wɨnɨkɨ-} & \obj{m-ehurka-} & \obj{o-wnɨkɨ-} & \obj{m-eɸɨrka-} & \obj{mɨ-wɨnk-} \\
\gl{1+2} & \rc{t-eɸurka-} & \rc{tɨt-wɨnɨkɨ-} & \obj{t-ehurka-} & \obj{tɨ-nɨkɨ-} & \obj{t͡ʃ-eɸɨrka-} & \obj{tɨt-wɨnk-} \\
\gl{3} & \rc{ɲ-eɸurka-} & \rc{nɨ-wɨnɨkɨ-} & \obj{ɲ-ehurka-} & \obj{nɨ-nɨkɨ-} & \obj{ɲ-eɸɨrka-} & \obj{nɨ-wɨnk-} \\
\end{tabular}\\
\parencites[510]{howard2001wrought}[189--191]{hixkaryanaderby1985}[209--211]{hawkins1953waiwai}[50]{waiwaihawkins1998}
\xe
%
\waiwai \obj{ka} \qu{to say} does not take \obj{kɨ-}, but rather conservative \obj{wɨ-} \exref{hixɨ.wai-108}.
Its \hixka counterpart also takes a reflex of \gl{1}\gl{s_a_} \rc{w-}, which has the form \obj{ɨ-} \exref{hixɨ.hix-118}.
A formally identical prefix is found on transitive verbs in \gl{1}>\gl{3} scenarios in \hixka \exref{hixɨ.hix-17}, where it also corresponds to \waiwai \obj{w(ɨ)-} \exref{hixɨ.wai-159}.

\pex<hixɨ>
\a<wai-108> \waiwai \parencite[][71]{waiwaihawkins1998}\\
\begingl
\glpreamble wɨɨkekɲe//
\gla wɨ-ka-jakɲe//
\glb \gl{1}-say-\gl{pst}//
\glft \qu{I said.}//
\endgl
\a<hix-118> \hixka \parencite[][124]{hixkaryanaderby1985}\\
\begingl
\glpreamble roxehra nay hamɨ Kaywerye ɨkekonɨ//
\gla ro-ʃe-hɨra n-a-je hamɨ kajwerʲe ɨ-ka-jakonɨ//
\glb \gl{1}-\gl{des}-\gl{neg} \gl{3}-be-\gl{npst}.\gl{uncert} \gl{evid} K. \gl{1}\gl{s_a_}-say-\gl{rem}.\gl{cont}//
\glft \qu{I said (to myself), “Kaywerye evidently doesn't like me”.}//
\endgl
\a<hix-17> \hixka \parencite[][191]{hixkaryanaderby1985}\\
\begingl
\gla ɨ-koroka-no//
\glb \gl{1}>\gl{3}-wash-\gl{imm}//
\glft \qu{I washed him.}//
\endgl
\a<wai-159> \waiwai \parencite[][192]{waiwaihawkins1998}\\
\begingl
\glpreamble wîyesî//
\gla wɨ-jo-jasɨ//
\glb \gl{1}>\gl{3}-boil-\gl{npst}//
\glft \qu{I will boil it.}//
\endgl
\xe
%
This correspondence allows us to establish \hixka \obj{ɨ-} a reflex of \rc{wɨ-}, similar to \ikpeng \pcref{sec:pekodian}.
Notably, \textcite{hixkaryanaderby1985} does not see this \obj{ɨ-} as an irregular \gl{1}\gl{s_a_}, but as the regular \gl{1}>\gl{3} marker, because he considers \hixka \obj{ka} \qu{to say} to be transitive, an issue which will be discussed in \cref{sec:say}.

There are three more verbs which did not take innovative \rc{k-} in \PWai, shown alongside \rc{ka} \qu{to say} in \cref{tab:pwaiverbs}.
While the two roots for \qu{to be} are straightforwardly reconstructible, \qu{to go} is somewhat of a special case.
While \hixka has the expected \obj{ɨ-}, \waiwai seems to have combined innovative \gl{1}\gl{s_a_} \obj{k-} with the old \rc{w-}, an analysis also considered by \textcite[90]{gildea1998}.
Alternatively, this form was influenced by deverbalized forms of \qu{to go}, where a reflex of the \gl{s_a_} class marker \rc{w-} has become fossilized \exref{waiwto}.

\pex<waiwto> \waiwai reflexes of the \gl{s_a_} class marker \rc{w-}
\a \obj{o-\emp{w}to-topo-nho} \qu{my trip} \parencite[92]{waiwaihawkins1998}
\a \obj{o-\emp{w}to-t͡ʃhe} \qu{after I went} \parencite[165]{waiwaihawkins1998}
\a \obj{kɨ-\emp{w}to-me} \qu{for us to go} \parencite[204]{waiwaihawkins1998}
\xe
%
In any case, \hixka \qu{to go} was clearly not affected by the extension of \rc{k-}, allowing us to reconstruct a \PWai first person form \rc{wɨ-tom-}.

@@

Summing up, we reconstruct the four verbs \rc{et͡ʃi} and \rc{a(h)} \qu{to be}, \rc{ka(s)} \qu{to say}, and \rc{tom} \qu{to go} as preserving the old \gl{1}\gl{s_a_} marker \rc{w-} in \PWai, while the rest took on innovative \rc{k-}.
